### 2010-10-24 Thesis/Chapter3.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{A practical approach of the finite element method}
\label{chap3}
\begin{shortAbstract}
The previous chapter explained in details how physics of anatomical structures could be described by continuun mechanics. When an organ is submitted to a force, the first effect of this application is a change in the organ's shape. This deformation can be measured using different strain tensors. By modelling the mechanical aspects of the material with an appropriate constitutive law, we can then deduce the stress tensor (or internal forces) at each point within the organ. The next step is to apply Newton's second law (law of conservation of momentum), which yields a partial differential equation at every point of the volume. The finite element method is a numerical procedure that allows to turn an infinite number of partial differential equations into a finite number of algebraic equations. The method follows different steps. We will first begin by describing the subdivision of the domain into smaller sub-domains into which local simpler equations are derived. We will then discuss how all the local equations are summed over the whole domain which eventually leads to the global solution. 
\end{shortAbstract}

@@

It is crucial to keep in mind that approximations occur at different stages during finite element analysis. The division of the whole domain into finite elements may not be exact (see \fig{chap3:fig-errorWithSubdivision}), introducing error in the domain being modelled. The second stage is when element equations are derived. As mentionned earlier, the unknowns of the problem are approximated using the idea that any continuous function can be representated by a linear combination of known functions and undetermined coefficients. Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element. There are a few types of approach for establishing these equations but, without going into details, the mathematical foundations of all these approaches are energy principles or the \emph{weighted residual methods} (see Appendix~\ref{appendix1}), which both lead to integrals during the process. Therefore, the second stage creates two sources of error: the representation of the solution by a linear combination of functions and the evaluation of the integrals. Finally, errors are introduced when solving the assembled system of algebraic equations. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision1.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision2.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision3.pdf}}
\caption[Error introduced by the division into elements]{The division of the domain (a) into elements is not always perfect (b). The discretisation error may be reduced by using smaller elements (c). However, this improvement is obtained to the detriment of computational efficiency as the number of algebraic equations to solve will increase.}
\label{chap3:fig-errorWithSubdivision}
\end{figure}

@@

The use of the natural coordinate system has several advantages \citep{Biswas76}. Not only the interpolation functions can be derivated only once per type of element in the mesh, but their expression is also much simpler, regardless of the actual element shape. Consequently, the element equations and the derived element matrices get to be simplified too. However, we now need to switch between the natural and global coordinate systems before solving the equations on the whole domain. Indeed, it is necessary to account for the diversity of element shapes within the mesh. Solving the element equations expressed into the natural coordinate system would not yield the expected result for the whole domain.

@@

\noindent \textbf{Remark.}
It is worth noting that \eqref{chap3:shapeFunctions} can be easily extended to three-dimensional problems. Let us consider a Cartesian coordinate system and we denote the position of node $ i $ $ \mathbf{x}_i = [x_{i1}, x_{i2}, x_{i3}]^T$. The position of a point $ \mathbf{x} = [x_{1}, x_{2}, x_{3}]^T$ within the element may then be interpolated by applying \eqref{chap3:shapeFunctions} to each component of the position:
\begin{align}
x_1 &= \sum_{i=1}^m x_{i1} \hat{\psi}_i^e(\xi) \notag \\
x_2 &= \sum_{i=1}^m x_{i2} \hat{\psi}_i^e(\xi) \\
x_3 &= \sum_{i=1}^m x_{i3} \hat{\psi}_i^e(\xi) \notag
\end{align}
If we construct the vector $ \hat{\mathbf{x}}^e $ of all nodal positions $ \mathbf{x}_i $ of the element, we can define a shape function matrix $ \mathbf{H} $ such as:
\begin{equation}
\label{chap3:H}
\mathbf{x} = \mathbf{H} \hat{\mathbf{x}}^e
\end{equation}
where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from submatrices $ \mathbf{H}_i $
\begin{equation}
\mathbf{H}_i = 
	\begin{bmatrix}
		\hat{\psi}_i^e & 0 & 0 \\
		0 & \hat{\psi}_i^e & 0 \\
		0 & 0 & \hat{\psi}_i^e
	\end{bmatrix}
\quad i = 1, 2, \ldots, m.
\end{equation}

@@

Consequently, a weak form is often created using energy principles or weighted residual methods. The weak form is frequently an integral form and requires a weaker continuity on the field variables. Hence, a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry. For this reason, the weak form is usually preferred over the strong form for obtaining an approximate solution of a practical engineering problem. \OFF

@@

The assembly operation is actually quite simple. Let us consider an example with stiffness matrices for an one-dimensional problem. We assume that we have three two-node elements. And we suppose the element connectivity information given by Table \ref{chap3:connectivity}, which provides the global node number for each local node number within an element. 
%
\begin{table}[ht]
	\begin{center}
	\begin{tabular}{|c|c|c|}
		\hline Elements & 1 & 2 \\ 
		\hline 1 & 1 & 2 \\ 
		\hline 2 & 2 & 3 \\ 
		\hline 3 & 3 & 4 \\ 
		\hline 
		\end{tabular} \caption{Element connectivity information}
		\label{chap3:connectivity}
	\end{center}
\end{table}
%
We also assume the following stiffness matrices for the three elements:
\begin{equation}
\mathbf{K}^1 = 
\color{blue}
	\begin{bmatrix}
	-1 & 2 \\
	3 & 1
	\end{bmatrix}
\color{black}
\text{, } \quad 
\mathbf{K}^2 = 
\color{green}
	\begin{bmatrix}
	5 & 2 \\
	-2 & 1
	\end{bmatrix}
\color{black}
\text{ and } \quad 
\mathbf{K}^3 = 
\color{red}
	\begin{bmatrix}
	3 & 1 \\
	2 & -1
	\end{bmatrix}
\color{black}
\end{equation}
The assembled matrix is a square matrix of dimension number of nodes $\times$ number of degrees of freedom. In our case, the global matrix is therefore a $ 4 \times 4 $ and is assembled in the following manner:
\begin{equation}
\mathbf{K} = 
	\begin{bmatrix}
	\color{blue}{-1} & \color{blue}{2} & \color{black}{0} & 0 \\
	\color{blue}{3} & \color{blue}{1}\color{black}{+}\color{green}{5} & \color{green}{2} & \color{black}{0} \\
	0 & \color{green}{-2} & \color{green}{1} \color{black}{+} \color{red}{3} & \color{red}{1} \\
	\color{black}{0} & 0 & \color{red}{2} & \color{red}{-1}
	\end{bmatrix}
\color{black}
\end{equation}
In other words, the global matrix is merely a rewritting on the whole domain of the element stiffness matrices, which are expressed with respect to local node numbers, in terms of global node numbers. In addition to constructing the global mass and damping matrices (as shown in section \ref{chap3:wordOnMatrices}), this process leads to the following global system of equations:
\begin{equation}
\label{chap3:dynamic2}
\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
\end{equation}	
The same set of equations was established at the element level in section \ref{chap3:dynamic} but we may note that the same system is also valid for the whole domain.

@@

\ON
\paragraph*{Cholesky decomposition.}	\label{chap3:CholeskyDecompo}
Cholesky decomposition allows to write a symmetric and definite positive matrix as the product of a lower triangular matrix with strictly positive diagonal entries by the transpose: 
\begin{equation}
\label{chap3:Cholesky}
\mathbf{A} = \mathbf{L} \mathbf{L}^T,
\end{equation}
where $ \mathbf{L} $ is lower triangular with strictly positive diagonal entries. In general, Cholesky decompositions are not unique. Similarly to LU decomposition, a Cholesky decomposition may be used to break one set of equations into two successive but simpler systems. 
\OFF

@@

\ON
\paragraph*{Pre-conditioners}
Unfortunately, it is common for the matrix $ \mathbf{A} $ not to be well conditioned. Consequently, the convergence of the chosen iterative method is often slow. In order to achieve an accelerated convergence, we may use a \emph{pre conditioner}. Although the formation and the study of pre-conditioners are out of the scope of this PhD thesis, pre-conditioners are a key element for solving systems of equations. To be effective, a preconditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
\begin{equation}
\label{chap3:preconditionedSystem}
\mathbf{P}^{-1} \mathbf{A} \mathbf{x} = \mathbf{P}^{-1} \mathbf{b}.
\end{equation}
Two criteria should also be considered when forming a preconditioner: (1) the ease with which it may be evaluated and (2) its memory requirements. The easiest pre-conditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case. Of course, such a conditioner leads to the resolution of the original system and is therefore totally useless. In constrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration. In practice, an optimal pre-conditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the pre-conditioner simple. Finding a good pre-conditioner for all types of problems is difficult. A pre-conditioner is only useful if the overall time spent on both the formation of  $ \mathbf{P} $ and the resolution of \eqref{chap3:preconditionedSystem} is less than the time spent at solving \eqref{chap3:linearEquation} alone.

One of the simplest pre-conditioner is the \emph{Jacobi pre-conditioner} where the pre-conditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $. This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant. Sparse approximations of LU or Cholesky decompositions may also be used as pre-conditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details). 
\OFF

>>>>

\chapter{A practical approach of the finite element method}
\label{chap3}
\begin{shortAbstract}
The previous chapter explained in details how physics of anatomical structures could be described by continuum mechanics. When an organ is submitted to a force, the first effect of this application is a change in the organ's shape. This deformation can be measured using different strain tensors. By modelling the mechanical aspects of the material with an appropriate constitutive law, we can then deduce the stress tensor (or internal forces) at each point within the organ. The next step is to apply Newton's second law (law of conservation of momentum), which yields a partial differential equation at every point of the volume. The finite element method is a numerical procedure that allows to turn an infinite number of partial differential equations into a finite number of algebraic equations. The method follows different steps. We will first begin by describing the subdivision of the domain into smaller sub-domains into which local simpler equations are derived. We will then discuss how all the local equations are summed over the whole domain which eventually leads to the global solution. 
\end{shortAbstract}

@@

It is crucial to keep in mind that approximations occur at different stages during finite element analysis. The division of the whole domain into finite elements may not be exact (see \fig{chap3:fig-errorWithSubdivision}), introducing error in the domain being modelled. The second stage is when element equations are derived. As mentioned earlier, the unknowns of the problem are approximated using the idea that any continuous function can be represented by a linear combination of known functions and undetermined coefficients. Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element. There are a few types of approach for establishing these equations but, without going into details, the mathematical foundations of all these approaches are energy principles or the \emph{weighted residual methods} (see Appendix~\ref{appendix1}), which both lead to integrals during the process. Therefore, the second stage creates two sources of error: the representation of the solution by a linear combination of functions and the evaluation of the integrals. Finally, errors are introduced when solving the assembled system of algebraic equations. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision1.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision2.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision3.pdf}}
\caption[Error introduced by the division into elements]{The division of the domain (a) into elements is not always perfect (b). The discretisation error may be reduced by using smaller elements (c). However, this improvement is obtained to the detriment of computational efficiency as the number of algebraic equations to solve will increase.}
\label{chap3:fig-errorWithSubdivision}
\end{figure}

@@

The use of the natural coordinate system has several advantages \citep{Biswas76}. Not only the interpolation functions can be derived only once per type of element in the mesh, but their expression is also much simpler, regardless of the actual element shape. Consequently, the element equations and the derived element matrices get to be simplified too. However, we now need to switch between the natural and global coordinate systems before solving the equations on the whole domain. Indeed, it is necessary to account for the diversity of element shapes within the mesh. Solving the element equations expressed into the natural coordinate system would not yield the expected result for the whole domain.

@@

\noindent \textbf{Remark.}
It is worth noting that \eqref{chap3:shapeFunctions} can be easily extended to three-dimensional problems. Let us consider a Cartesian coordinate system and we denote the position of node $ i $ $ \mathbf{x}_i = [x_{i1}, x_{i2}, x_{i3}]^T$. The position of a point $ \mathbf{x} = [x_{1}, x_{2}, x_{3}]^T$ within the element may then be interpolated by applying \eqref{chap3:shapeFunctions} to each component of the position:
\begin{align}
x_1 &= \sum_{i=1}^m x_{i1} \hat{\psi}_i^e(\xi) \notag \\
x_2 &= \sum_{i=1}^m x_{i2} \hat{\psi}_i^e(\xi) \\
x_3 &= \sum_{i=1}^m x_{i3} \hat{\psi}_i^e(\xi) \notag
\end{align}
If we construct the vector $ \hat{\mathbf{x}}^e $ of all nodal positions $ \mathbf{x}_i $ of the element, we can define a shape function matrix $ \mathbf{H} $ such as:
\begin{equation}
\label{chap3:H}
\mathbf{x} = \mathbf{H} \hat{\mathbf{x}}^e
\end{equation}
where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from sub-matrices $ \mathbf{H}_i $
\begin{equation}
\mathbf{H}_i = 
	\begin{bmatrix}
		\hat{\psi}_i^e & 0 & 0 \\
		0 & \hat{\psi}_i^e & 0 \\
		0 & 0 & \hat{\psi}_i^e
	\end{bmatrix}
\quad i = 1, 2, \ldots, m.
\end{equation}

@@

Consequently, a weak form is often created using energy principles or weighted residual methods. The weak form is frequently an integral form and requires a weaker continuity on the field variables. Hence, a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry. For this reason, the weak form is usually preferred over the strong form for obtaining an approximate solution of a practical engineering problem.

@@

The assembly operation is actually quite simple. Let us consider an example with stiffness matrices for an one-dimensional problem. We assume that we have three two-node elements. And we suppose the element connectivity information given by Table \ref{chap3:connectivity}, which provides the global node number for each local node number within an element. 
%
\begin{table}[ht]
	\begin{center}
	\begin{tabular}{|c|c|c|}
		\hline Elements & 1 & 2 \\ 
		\hline 1 & 1 & 2 \\ 
		\hline 2 & 2 & 3 \\ 
		\hline 3 & 3 & 4 \\ 
		\hline 
		\end{tabular} \caption{Element connectivity information}
		\label{chap3:connectivity}
	\end{center}
\end{table}
%
We also assume the following stiffness matrices for the three elements:
\begin{equation}
\mathbf{K}^1 = 
\color{blue}
	\begin{bmatrix}
	-1 & 2 \\
	3 & 1
	\end{bmatrix}
\color{black}
\text{, } \quad 
\mathbf{K}^2 = 
\color{green}
	\begin{bmatrix}
	5 & 2 \\
	-2 & 1
	\end{bmatrix}
\color{black}
\text{ and } \quad 
\mathbf{K}^3 = 
\color{red}
	\begin{bmatrix}
	3 & 1 \\
	2 & -1
	\end{bmatrix}
\color{black}
\end{equation}
The assembled matrix is a square matrix of dimension number of nodes $\times$ number of degrees of freedom. In our case, the global matrix is therefore a $ 4 \times 4 $ and is assembled in the following manner:
\begin{equation}
\mathbf{K} = 
	\begin{bmatrix}
	\color{blue}{-1} & \color{blue}{2} & \color{black}{0} & 0 \\
	\color{blue}{3} & \color{blue}{1}\color{black}{+}\color{green}{5} & \color{green}{2} & \color{black}{0} \\
	0 & \color{green}{-2} & \color{green}{1} \color{black}{+} \color{red}{3} & \color{red}{1} \\
	\color{black}{0} & 0 & \color{red}{2} & \color{red}{-1}
	\end{bmatrix}
\color{black}
\end{equation}
In other words, the global matrix is merely a rewriting on the whole domain of the element stiffness matrices, which are expressed with respect to local node numbers, in terms of global node numbers. In addition to constructing the global mass and damping matrices (as shown in section \ref{chap3:wordOnMatrices}), this process leads to the following global system of equations:
\begin{equation}
\label{chap3:dynamic2}
\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
\end{equation}	
The same set of equations was established at the element level in section \ref{chap3:dynamic} but we may note that the same system is also valid for the whole domain.

@@

\paragraph*{Cholesky decomposition.}	\label{chap3:CholeskyDecompo}
Cholesky decomposition allows to write a symmetric and definite positive matrix as the product of a lower triangular matrix with strictly positive diagonal entries by the transpose: 
\begin{equation}
\label{chap3:Cholesky}
\mathbf{A} = \mathbf{L} \mathbf{L}^T,
\end{equation}
where $ \mathbf{L} $ is lower triangular with strictly positive diagonal entries. In general, Cholesky decompositions are not unique. Similarly to LU decomposition, a Cholesky decomposition may be used to break one set of equations into two successive but simpler systems.

@@

\paragraph*{Pre-conditioners.}
Unfortunately, it is common for the matrix $ \mathbf{A} $ not to be well conditioned. Consequently, the convergence of the chosen iterative method is often slow. In order to achieve an accelerated convergence, we may use a \emph{pre conditioner}. Although the formation and the study of pre-conditioners are out of the scope of this PhD thesis, pre-conditioners are a key element for solving systems of equations. To be effective, a pre conditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
\begin{equation}
\label{chap3:preconditionedSystem}
\mathbf{P}^{-1} \mathbf{A} \mathbf{x} = \mathbf{P}^{-1} \mathbf{b}.
\end{equation}
Two criteria should also be considered when forming a pre conditioner: (1) the ease with which it may be evaluated and (2) its memory requirements. The easiest pre-conditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case. Of course, such a conditioner leads to the resolution of the original system and is therefore totally useless. In contrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration. In practice, an optimal pre-conditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the pre-conditioner simple. Finding a good pre-conditioner for all types of problems is difficult. A pre-conditioner is only useful if the overall time spent on both the formation of  $ \mathbf{P} $ and the resolution of \eqref{chap3:preconditionedSystem} is less than the time spent at solving \eqref{chap3:linearEquation} alone.

One of the simplest pre-conditioner is the \emph{Jacobi pre-conditioner} where the pre-conditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $. This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant. Sparse approximations of LU or Cholesky decompositions may also be used as pre-conditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details).
